\documentclass[11pt]{article}

% ---------------------------------------------------------------
% NOTE TO AUTHORS
% Unified draft: the SNN framework (Bayesian dropout + subjective
% logic) is the spine; its uncertainty semantics are made
% principled via the law of total variance; the frozen-backbone
% "Subjective Head" is a deployment mode that enables strong
% backbones on modest hardware and the synthetic->real move.
% Result cells marked [fill] need numbers from runs (see code).
% Citations marked %VERIFY were added in the merge.
% Plain article class for portability; drop into the UAI style.
% ---------------------------------------------------------------

\usepackage[utf8]{inputenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}

\newtheorem{proposition}{Proposition}
\newcommand{\todo}[1]{\textcolor{red}{[#1]}}
\newcommand{\Eb}{\mathbb{E}}
\newcommand{\KL}{\mathrm{KL}}
\newcommand{\Var}{\operatorname{Var}}
\newcommand{\Cov}{\operatorname{Cov}}
\newcommand{\diag}{\operatorname{diag}}
\newcommand{\R}{\mathbb{R}}

\title{Subjective Neural Networks:\\
Trust-Aware Bayesian Dropout with a Principled Uncertainty Decomposition}
\author{Koffi Ismael Ouattara \quad Theo Dimitrakos \quad Frank Kargl}
\date{}

\begin{document}
\maketitle

\begin{abstract}
Neural networks achieve strong predictive accuracy but communicate uncertainty
poorly, which limits their use in safety-critical settings. Bayesian dropout
approximates a posterior predictive but conflates ignorance with stochastic
variability, while Evidential Deep Learning (EDL) imposes a Dirichlet opinion at the
output without a grounded posterior. We introduce the \textbf{Subjective Neural
Network (SNN)}, which unifies Bayesian variational inference with Subjective Logic by
placing a Beta-distributed \emph{trust} variable on each unit's participation via
Beta--Bernoulli dropout, sampling Bernoulli masks, and aggregating the posterior
predictive samples into a Dirichlet opinion with explicit belief and residual
uncertainty. Beyond the construction, we make the SNN's uncertainty \emph{semantics}
principled. Using the law of total variance---established machinery we apply rather
than claim---we show analytically that the SL uncertainty mass $u$ tracks only the
\emph{epistemic} (across-sample) term of the predictive variance, whereas the
predictive entropy $H$ and $1-\max_k b_k$ read the \emph{total} spread of the mean
prediction. This corrects the common reading of $u$ as a general ``I do not know''
signal and predicts, in particular, that a confident SNN reports low $u$ on
out-of-distribution (OOD) inputs. To evaluate on strong backbones with modest
hardware, we deploy the SNN as a \emph{post-hoc subjective head} on a frozen
foundation model: the backbone runs once and nested sampling is restricted to a small
bottleneck, making inference overhead negligible. On real foundation-model features we
(i) confirm the decomposition empirically, (ii) show a two-dimensional
$(\max_k b_k, u)$ abstention frontier that separates within-model instability from
lack of decisive evidence, and (iii) show trust-discounted SL fusion preserves accuracy
when an ensemble contains adversarially miscalibrated sources, where logit averaging
and Laplace ensembling collapse. We report modern UQ baselines, multiple seeds, full
scalar metrics, and release code.
\end{abstract}

% ===============================================================
\section{Introduction}
% ===============================================================
Neural networks (NNs) reach state-of-the-art accuracy across vision and language, yet
their inability to communicate \emph{meaningful} uncertainty remains a bottleneck for
deployment in healthcare, autonomous driving, and security. Overconfidence on incorrect
predictions erodes trust precisely where reliability matters most.

Two families address this. Bayesian NNs \citep{neal1996,blundell2015} and
dropout-as-Bayesian-inference \citep{gal2016b} approximate a posterior predictive and
capture model uncertainty, but express it only as variability in probability estimates
and do not separate committed belief from residual ignorance. Evidential approaches---
EDL \citep{sensoy2018}---instead output a Dirichlet at the head, yielding a Subjective
Logic opinion in a single forward pass, but recent work \citep{shen2024} shows EDL's
uncertainty need not decrease with data and can behave like an energy-based OOD score
rather than a grounded posterior. Thus Bayesian methods give posterior uncertainty
without structured trust semantics, while evidential methods give structure without a
grounded posterior. This is the structural gap our framework targets.

\paragraph{Subjective Neural Networks.}
We propose the \textbf{SNN}, which bridges Bayesian dropout and Subjective Logic
\citep[SL;][]{josang2016}. Each unit $j$ carries a latent trust probability
$p_j\sim\mathrm{Beta}(\alpha_j,\beta_j)$ governing its participation through a Bernoulli
mask $z_j\sim\mathrm{Bernoulli}(p_j)$; the known Beta--binomial-opinion mapping lets us
read $p_j$ as a subjective trust in unit $j$. A nested sampling procedure (sample trust,
sample masks, aggregate) produces posterior-predictive samples that we project onto a
Dirichlet, yielding a multinomial opinion $\omega=(b,u,a)$ with an explicit
belief/ignorance split. Unlike EDL, the Dirichlet \emph{emerges} from posterior
predictive sampling rather than being imposed at the output.

\paragraph{Making the motivation principled.}
The promise attached to such constructions is that the uncertainty mass $u$ measures
epistemic uncertainty. We show this reading is only half correct, and we make precise
which half, using the law of total variance \citep{kwon2020,depeweg2018}---which we
treat as known machinery, not a contribution. The SNN's predictive variance decomposes
into an aleatoric term (expected within-model categorical spread) and an epistemic term
(variance of the mean prediction across the trust/mask posterior). By construction the
moment-matched $u$ is a monotone function of the \emph{epistemic} term alone and is
blind to aleatoric spread in the mean. This yields a falsifiable prediction: a
confident SNN reports \emph{low} $u$ on OOD inputs that excite familiar pathways,
because masks do not move a confident-but-wrong prediction; $H$ and $1-\max_k b_k$,
which read the mean prediction, do detect them. We confirm this on real features
(\S\ref{sec:exp}). The decomposition is the principled spine: it gives each SNN signal a
precise meaning and explains, rather than asserts, the model's behaviour.

\paragraph{Strong backbones on modest hardware.}
Bayesian dropout over an entire deep network is the costly regime that motivated the
``efficiency'' concerns of prior critiques. We instead deploy the SNN as a
\emph{post-hoc subjective head} on a \emph{frozen} foundation model: the SNN's
Beta--Bernoulli dropout acts only on a small bottleneck, the backbone runs once, and
inference overhead is well under one percent of backbone compute. This (a) replaces the
underpowered LeNet/ResNet-18 backbones of the original evaluation with modern
representations, (b) is trainable on a laptop because only the head is learned, and
(c) takes the analysis from synthetic to real data.

\paragraph{Contributions.}
\begin{enumerate}
\item The \textbf{SNN}: a Bayesian-dropout framework with neuron-level Beta trust that
produces SL opinions from posterior-predictive sampling, with an \emph{explicit}
discrete-mask gradient path (Concrete relaxation) and an account of the variational
loss-balance pitfall that silently disables the mechanism in small heads.
\item A \textbf{principled reading} of the SNN's uncertainty via the law of total
variance: $u$ is the epistemic (across-sample) term; $H$ and $1-\max_k b_k$ read the
total spread of the mean. We derive analytically why $u$ is not a general OOD signal.
\item A \textbf{frozen-backbone deployment} of the SNN with negligible overhead,
enabling evaluation on real foundation-model features with strong backbones.
\item \textbf{Operational results}: a $(\max_k b_k, u)$ two-threshold abstention
frontier that significantly beats each individual SL component in extrapolation regimes,
and trust-discounted SL fusion that survives adversarially miscalibrated sources where
logit averaging and Laplace ensembles collapse. Evaluated with modern baselines,
multiple seeds with CIs, and full scalar metrics; code released.
\end{enumerate}

% ===============================================================
\section{Background and Related Work}
% ===============================================================
\paragraph{Uncertainty in deep learning.}
BNNs \citep{neal1996,blundell2015} model weight distributions but are costly; MC Dropout
\citep{gal2016b} is a scalable approximation; variational and Concrete dropout
\citep{molchanov2017,gal2017concrete,maddison2017} learn or relax dropout rates. All
express uncertainty as predictive variance without a belief/ignorance split.
Beta--Bernoulli dropout \citep{lee2019} places a Beta prior over keep probabilities for
\emph{sparsification}; we repurpose its variational machinery for trust-aware
uncertainty.

\paragraph{Subjective Logic and the Dirichlet correspondence.}
An SL multinomial opinion is $\omega=(b,u,a)$ with $\sum_k b_k+u=1$, mapping to a
Dirichlet with concentration $\eta$:
\begin{equation}
b_k=\frac{\eta_k}{S+K},\qquad u=\frac{K}{S+K},\qquad a_k=\tfrac1K,\qquad S=\textstyle\sum_k\eta_k,
\end{equation}
and projected probability $p_k=b_k+u\,a_k$ \citep{josang2016}. SL has been used to assess
dataset and network trustworthiness \citep{cheng2020,ouattara2025a,ouattara2025b}.

\paragraph{Evidential Deep Learning.}
EDL \citep{sensoy2018} outputs evidence $e_k$, sets $\eta_k=e_k+1$, trains with an SSE
loss plus a KL regulariser, and produces an opinion in one pass. \citet{shen2024}
question whether its $u$ is a coherent posterior signal.

\paragraph{Last-layer Laplace and ensembles.}
Last-layer Laplace \citep{daxberger2021} is a cheap, post-hoc Gaussian over the final
linear layer; Deep Ensembles \citep{lakshminarayanan2017} are the standard strong UQ
baseline; Posterior Networks \citep{charpentier2020} construct epistemic uncertainty
that provably decreases with data.

\paragraph{SL fusion operators.}
SL provides closed-form operators on $(b,u)$ \citep{josang2016}: cumulative fusion
$\eta^C=\sum_i\eta_i-(N-1)$, averaging $\eta^C=(\sum_i\eta_i)/N$, and trust transitivity
$b'_k=t\,b_k,\;u'=t\,u+(1-t)$, which carry source-reliability uncertainty through to the
fused opinion.

\paragraph{Uncertainty decomposition.}
The law of total variance splits predictive variance into expected conditional variance
(aleatoric) and variance of the conditional mean (epistemic) \citep{depeweg2018};
\citet{kwon2020} give the categorical form we use. We treat this as established.

% ===============================================================
\section{Variational Inference for the SNN}
% ===============================================================
The SNN is Bayesian over per-unit trust probabilities. Given data $\mathcal D$, exact
posterior inference is intractable, so we use variational inference, maximising the
evidence lower bound (ELBO)
\begin{equation}
\mathcal L(\phi)=\Eb_{q_\phi}[\log p(\mathcal D\mid\cdot)]-\KL\big(q_\phi\,\|\,p\big),
\end{equation}
optimised by stochastic gradient variational Bayes \citep{kingma2022,rezende2014} with
differentiable Monte Carlo estimators. This requires a reparameterisation of the latent.

\paragraph{Beta latents via Kumaraswamy.}
The Beta distribution lacks a convenient location--scale reparameterisation in its shape
parameters \citep{nalisnick2017}. We use the Kumaraswamy surrogate \citep{mitnik2013},
which admits a closed-form inverse CDF:
\begin{equation}
p_j=\big(1-u^{1/\beta_j}\big)^{1/\alpha_j},\qquad u\sim\mathrm{Uniform}(0,1),
\label{eq:kuma}
\end{equation}
giving differentiable samples of $p_j$ w.r.t.\ $(\alpha_j,\beta_j)$. We quantify the
Kumaraswamy--Beta gap in Appendix~\ref{app:impl}.

\paragraph{Discrete mask gradient (explicit).}
Conditioned on $p_j$, the mask $z_j$ is binary, so we relax it \emph{at training time}
with the Concrete/Gumbel-softmax distribution \citep{maddison2017}:
\begin{equation}
z_j=\sigma\!\Big(\big(\log p_j-\log(1-p_j)+\log u-\log(1-u)\big)/\tau\Big),\quad
u\sim\mathrm{Uniform}(0,1),
\label{eq:concrete}
\end{equation}
with temperature $\tau$; gradients flow through both $p_j$ (via \eqref{eq:kuma}) and
$z_j$ (via \eqref{eq:concrete}). At inference we use hard Bernoulli draws. This states
explicitly the gradient path that prior versions left unspecified.

% ===============================================================
\section{The Subjective Neural Network}
% ===============================================================
\paragraph{Model.}
For each unit $j$, $p_j\sim\mathrm{Beta}(\alpha_j,\beta_j)$ is a subjective trust in the
unit's activation, and $z_j\sim\mathrm{Bernoulli}(p_j)$ masks it. The forward pass under
a sampled mask is $f(x;W,z)=\mathrm{NN}(x;W\odot z)$. We are Bayesian over the
trust/keep probabilities, \emph{not} over weights; we state this rather than framing it
as weight-space BNN inference.

\paragraph{Training objective.}
With variational posterior $q(p_j)=\mathrm{Beta}(\alpha_j,\beta_j)$ and prior
$\mathrm{Beta}(a_0,b_0)$, the per-batch loss is the negative ELBO
\begin{equation}
\mathcal L=\frac{1}{|B|}\!\sum_{(x,y)\in B}\!-\log p(y\mid x,W\odot z)
+\frac{\beta(t)}{N_{\text{train}}}\sum_{j=1}^{H}
\KL\!\big(\mathrm{Beta}(\alpha_j,\beta_j)\,\|\,\mathrm{Beta}(a_0,b_0)\big),
\label{eq:elbo}
\end{equation}
with closed-form Beta--Beta KL (Appendix~\ref{app:impl}).

\paragraph{Loss-balance pitfall.}
For $H\!\in\![12,512]$ trust units and $N_{\text{train}}\!\sim\!10^3$--$10^5$, the
per-example KL is order $10^{-2}$ while the NLL is order $10^0$: the KL becomes rounding
error, the posterior collapses to a point, and the Beta--Bernoulli mechanism does
nothing. Three knobs fix this: a multiplier $\beta_{\max}\!\gg\!1$ tuned on validation
calibration (typically $[5,50]$); an informative prior peaked at a sensible keep
probability; and a KL warmup ramping $\beta(t)$ to $\beta_{\max}$. \emph{Diagnostic}: if
$\mathrm{std}_j(\Eb[p_j])<0.01$ after training, the mechanism is dead.

\paragraph{Inference: nested sampling and the augmented opinion.}
For input $x$, draw $N_p$ trust samples $p^{(i)}$; for each, $N_m$ masks $z^{(i,j)}$, and
compute $\pi^{(i,j)}=\mathrm{softmax}(f(x;W,z^{(i,j)}))$. Each trust sample yields a
per-sample Dirichlet from class counts, $\alpha^{(i)}_k=c^{(i)}_k+a_0$ with mean
$m^{(i)}=\alpha^{(i)}/S^{(i)}$. The law of total variance then splits the predictive
variance \emph{inside} the opinion construction:
\begin{equation}
\mathrm{varE}_k=\Var_i\big[m^{(i)}_k\big]\ \text{(epistemic)},\qquad
\mathrm{eVar}_k=\Eb_i\big[m^{(i)}_k(1-m^{(i)}_k)\big]\ \text{(aleatoric)}.
\end{equation}
The \textbf{augmented SL opinion} sets the vacuity to the \emph{epistemic share},
routes the aleatoric component into the Dirichlet concentration, and takes the
direction from the pooled evidence $\bar\alpha=\sum_i\alpha^{(i)}$:
\begin{equation}
u^\ast=\frac{\sum_k\mathrm{varE}_k}{\sum_k(\mathrm{varE}_k+\mathrm{eVar}_k)},\quad
b_k=(1-u^\ast)\frac{\bar\alpha_k}{\sum_j\bar\alpha_j},\quad
P_k=b_k+\frac{u^\ast}{K},\quad
S^\ast=\frac1K\sum_k\max\!\Big(1,\frac{P_k(1-P_k)}{\mathrm{eVar}_k}-1\Big),
\label{eq:augmented}
\end{equation}
with output Dirichlet $\alpha^\ast_k=P_kS^\ast$. Unlike a moment-matched/MLE
Dirichlet fit, whose $u$ is monotone in the epistemic variance alone
(Prop.~\ref{prop:u}) and collapses for a confident head, $u^\ast$ answers ``what
\emph{fraction} of my uncertainty is epistemic,'' and the aleatoric spread is
preserved in $S^\ast$ rather than discarded. Two cautions. (i) Defining
$\mathrm{eVar}$ as the per-sample Dirichlet \emph{posterior} variance is incorrect:
it and the multinomial noise in $\mathrm{varE}$ both shrink as $1/N_m$, so $u^\ast$
saturates near $1/2$ even in purely aleatoric regimes; the categorical form above
restores the separation. (ii) We use prior $a_0=1/K$ so the $+u^\ast/K$ base-rate
smoothing is not applied twice. For comparison we also report the legacy signals
$u=K/(S+K)$, $H$, and $1-\max_k b_k$ from the moment-matched fit
(eq.~\ref{eq:signals}).
\begin{equation}
u=\frac{K}{S+K},\qquad H=-\sum_k\bar\pi_k\log\bar\pi_k,\qquad \neg b=1-\max_k b_k.
\label{eq:signals}
\end{equation}

\paragraph{From unit-level trust to output uncertainty.}
The shape of each unit's Beta determines \emph{which kind} of uncertainty its
stochasticity injects into the output, and the nested sampler separates the two via
the LoTV split. With $p_j\sim\mathrm{Beta}(\alpha_j,\beta_j)$:
$\alpha_j\!\gg\!\beta_j$ (\emph{trusted}): the unit is essentially always active and
contributes no uncertainty; $\alpha_j\!\ll\!\beta_j$ (\emph{distrusted}): always
inactive, none either. $\alpha_j\!\approx\!\beta_j\!\gg\!1$ (\emph{aleatoric}): $p_j$
concentrates at $1/2$, so masks flip independently \emph{within} each trust draw---the
within-sample categorical variance ($\mathrm{eVar}$) carries the signal.
$\alpha_j\!\approx\!\beta_j\!\ll\!1$ (\emph{epistemic}): the U-shaped Beta makes
$p_j\!\approx\!0$ or $1$ \emph{per trust draw}, so predictions are consistent within a
draw but differ across draws---the across-sample variance ($\mathrm{varE}$) carries it.
The flat case $\alpha_j\!\approx\!\beta_j\!\approx\!1$ is mixed and, analytically,
aleatoric-leaning ($\Eb_p[g(1-g)]>\Var_p[g]$ for our worked example), so only the
U-shape is genuinely epistemic. We validate this on a minimal two-unit network with
deterministic priority inference ($P(0)=(1-p)p$, $P(1)=p$, $P(2)=(1-p)^2$): the
pooled counts for the aleatoric and epistemic regimes can be \emph{identical}
(e.g.\ $[0,1,2],[0,1,2],[0,1,2]$ versus $[0,0,0],[1,1,1],[2,2,2]$), so a pooled
Dirichlet cannot distinguish them; only the LoTV split inside the augmented opinion
can. Per-unit trust is itself a binomial SL opinion with $W{=}2$:
$u_j=2/(2+\alpha_j+\beta_j)$, $b_j=\alpha_j/(2+\alpha_j+\beta_j)$,
$d_j=\beta_j/(2+\alpha_j+\beta_j)$; sub-prior units ($\alpha_j,\beta_j<1$) correctly
receive \emph{higher} vacuity than $\alpha_j=\beta_j=1$. Reporting the learned regime
distribution per layer makes the ``trust-aware'' claim inspectable.

\paragraph{Deployment as a frozen-backbone head.}
The SNN need not be a full network. When the backbone $g$ is a frozen foundation model,
we apply Beta--Bernoulli dropout only in a two-layer head on the penultimate
representation $x=g(\text{input})$:
$x\!\xrightarrow{\text{LN}}\!\tilde x\!\xrightarrow{W_1,\mathrm{ReLU}}\!h\!\xrightarrow{\odot z}\!h'\!\xrightarrow{W_2}\!\text{logits}$.
The backbone runs once; nested sampling loops only over the small head. This is the same
SNN method restricted to the head, and it is what makes strong backbones laptop-feasible.

\begin{algorithm}[t]
\caption{SNN training (one minibatch)}
\begin{algorithmic}[1]
\State sample $u\sim\mathrm{Uniform}$; $p_j\!\leftarrow\!(1-u^{1/\beta_j})^{1/\alpha_j}$ \Comment{Kumaraswamy}
\State sample relaxed mask $z_j$ via Concrete \eqref{eq:concrete}
\State $\hat y\!\leftarrow\! f(x;W\odot z)$; compute $\mathcal L$ \eqref{eq:elbo} with KL warmup $\beta(t)$
\State update $\{\alpha_j,\beta_j\}$ (and head weights $W$) by backprop
\end{algorithmic}
\end{algorithm}

% ===============================================================
\section{Principled Uncertainty Semantics via the Law of Total Variance}
\label{sec:lotv}
% ===============================================================
Let $\omega=(p,z)$ be the SNN's latent and let the label be one-hot $y$ with
$y\mid\omega\sim\mathrm{Categorical}(\pi(x,\omega))$, so $\Eb[y\mid\omega]=\pi$ and
$\Cov(y\mid\omega)=\diag(\pi)-\pi\pi^\top$. The law of total covariance
\citep{kwon2020,depeweg2018} gives
\begin{equation}
\underbrace{\Cov[y\mid x]}_{\text{total}}
=\underbrace{\Eb_\omega\!\big[\diag(\pi)-\pi\pi^\top\big]}_{\text{aleatoric }A(x)}
+\underbrace{\Cov_\omega[\pi]}_{\text{epistemic }E(x)},
\label{eq:lotv}
\end{equation}
and, in trace form with $\bar\pi=\Eb_\omega[\pi]$,
\begin{equation}
\underbrace{1-\lVert\bar\pi\rVert^2}_{\text{total}}
=\underbrace{\Eb_\omega[1-\lVert\pi\rVert^2]}_{\text{aleatoric}}
+\underbrace{\Eb_\omega\lVert\pi\rVert^2-\lVert\bar\pi\rVert^2}_{\text{epistemic}} .
\label{eq:lotv-trace}
\end{equation}
The epistemic term is the across-sample variance of the softmax output; the aleatoric
term is the expected within-sample categorical spread. The SNN's nested sampler
estimates both directly:
$\widehat{\operatorname{tr}E}=\tfrac1N\sum_n\lVert\pi_n-\bar\pi\rVert^2$ and
$\widehat{\operatorname{tr}A}=\tfrac1N\sum_n(1-\lVert\pi_n\rVert^2)$.

\begin{proposition}[$u$ is an epistemic-only signal]
\label{prop:u}
With Dirichlet moment matching as in \S5, the per-class variance setting the total
evidence is $v_k=\Var_i[\bar\pi^{(i)}_k]$, the across-trust-sample (epistemic) variance
of the mean prediction. Since $S=\mathrm{median}_k[m_k(1-m_k)/v_k-1]$ is decreasing in
$v_k$ and $u=K/(S+K)$ is increasing in $1/S$, $u$ is a strictly increasing function of
the epistemic term and, holding $\bar\pi$ fixed, is invariant to the aleatoric spread of
the mean prediction.
\end{proposition}

\noindent\emph{Why $u$ fails on OOD.}
Let $h=\mathrm{ReLU}(W_1\tilde x+b_1)$ and $\ell=W_2(h\odot z)+b_2$, so
$\Var_z[\ell_k]=\sum_j w_{k,j}^2 h_j^2 p_j(1-p_j)$. When a trained head has a dominant
class $k^*$ for $h$, the logit gap dwarfs $(\Var_z[\ell_{k^*}])^{1/2}$ and $\pi$ is
near-constant across masks: $E(x)\!\to\!0$, $S\!\to\!\infty$, $u\!\to\!0$. This depends
only on $W_2,h$, not on whether $x$ is in-distribution; an OOD input exciting the same
pathway gives the same low $u$ while $\bar\pi$ is wrong and spread out (high $H$, low
$\max_k b_k$). The decomposition thus \emph{predicts} the OOD failure of $u$ and the
success of $H,\neg b$. The pair $(\max_k b_k, u)$ recovers both terms of \eqref{eq:lotv};
$H$ implicitly aggregates them. This is the principled correction to the naive reading
of $u$ as a general ignorance signal.

% ===============================================================
\section{Experiments}
\label{sec:exp}
% ===============================================================
\paragraph{Setup.}
Backbone: frozen \textbf{DINOv2 ViT-S/14} \citep{oquab2024} %VERIFY
(alt.\ CLIP ViT-B/32 \citep{radford2021}); %VERIFY
features extracted once. The SNN head and all baselines share the same
LN--Linear--ReLU--Linear architecture on the same frozen features, isolating the
uncertainty mechanism from backbone capacity. ID: CIFAR-10, CIFAR-100; OOD: SVHN and a
held-out CIFAR-100 split. Baselines: deterministic Linear, MC Dropout \citep{gal2016b},
EDL \citep{sensoy2018}, last-layer Laplace \citep{daxberger2021}, Deep Ensembles
\citep{lakshminarayanan2017}, optional Posterior Network \citep{charpentier2020}.
Metrics: Accuracy, NLL, Brier, ECE (15 bins), OOD AUROC/AUPR/FPR@95, selective AUSC,
over \textbf{10 seeds} with 95\% CIs. Robustness with FGSM \emph{and} PGD. We report
wall-clock inference time and peak memory for a deterministic pass, MC Dropout (100
passes), and the SNN head ($N_pN_m$ head passes + moment matching), substantiating the
negligible-overhead claim.

% ---------------------------------------------------------------
\subsection{Predictive performance, calibration, and OOD}
% ---------------------------------------------------------------
The SNN's original evaluation axes (accuracy, calibration, OOD), now on real features.
\begin{table}[h]\centering
\caption{Main comparison on frozen DINOv2 features. \todo{fill from runs}.}
\begin{tabular}{lcccccc}
\toprule
Method & Acc. & NLL & Brier & ECE & OOD AUROC & FPR@95 \\
\midrule
Linear & \todo{} & \todo{} & \todo{} & \todo{} & \todo{} & \todo{} \\
MC Dropout & \todo{} & \todo{} & \todo{} & \todo{} & \todo{} & \todo{} \\
EDL & \todo{} & \todo{} & \todo{} & \todo{} & \todo{} & \todo{} \\
Deep Ensemble & \todo{} & \todo{} & \todo{} & \todo{} & \todo{} & \todo{} \\
Laplace & \todo{} & \todo{} & \todo{} & \todo{} & \todo{} & \todo{} \\
SNN (ours) & \todo{} & \todo{} & \todo{} & \todo{} & \todo{($H$)} & \todo{} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Decomposition analysis}
On the same trained SNN, the three signals \eqref{eq:signals} are not interchangeable:
selective AUROC is similar, but OOD AUROC collapses for $u$ while $H,\neg b$ stay high,
confirming Proposition~\ref{prop:u}. We also plot the empirical
aleatoric/epistemic split \eqref{eq:lotv-trace} for ID vs OOD. \todo{fill}.

\subsection{Covariate shift (rotated inputs)}
Retained from the SNN evaluation: predictive class probabilities and uncertainty as
inputs are progressively rotated. Expected behaviour: SNN uncertainty rises smoothly
under shift, between the overconfident MC Dropout and the over-cautious EDL. \todo{figs}.

\subsection{Two-threshold abstention}
Multi-regime test mixture (easy / split / diffuse) with the head trained on easy to
force extrapolation. We sweep 1D rules ($H$, $\neg b$, $u$) and the 2D rule (accept iff
$\max_k b_k>\tau_b$ and $u<\tau_u$) and integrate the retained-accuracy/coverage Pareto
envelope.
\begin{table}[h]\centering
\caption{Selective-prediction AUSC, 10 seeds, train-on-easy. \todo{fill}; expected:
2D $>$ each SL component ($p<0.01$); not significantly above $H$.}
\begin{tabular}{lcc}
\toprule
Signal & Mean AUSC (95\% CI) & $p$ vs.\ 2D \\
\midrule
$H$ & \todo{} & \todo{} \\
$\neg b$ & \todo{} & \todo{} \\
$u$ & \todo{} & \todo{} \\
$(\max b,u)$ 2D & \todo{} & ref. \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Adversarial-robust SL fusion}
Five SNN sources; two reliable, three adversarial (labels rotated $y'=(y+1)\bmod K$).
Trust-discounted SL fusion uses $t_i=(n_{\text{correct},i}+1)/(n_{\text{total},i}+2)$.
\begin{table}[h]\centering
\caption{Fusion across 2 reliable + 3 adversarial sources. \todo{fill}; expected: logit
averaging and Laplace ensemble collapse; trust-discounted SL preserves accuracy and
returns a full opinion with elevated $u$.}
\begin{tabular}{lccc}
\toprule
Method & Acc. & NLL & ECE \\
\midrule
Single best source & \todo{} & \todo{} & \todo{} \\
Logit averaging & \todo{} & \todo{} & \todo{} \\
Acc-weighted vote & \todo{} & \todo{} & \todo{} \\
Laplace ensemble & \todo{} & \todo{} & \todo{} \\
SL averaging / cumulative & \todo{} & \todo{} & \todo{} \\
SL trust-discounted & \todo{} & \todo{} & \todo{} \\
\bottomrule
\end{tabular}
\end{table}

% ===============================================================
\section{Discussion}
% ===============================================================
\textbf{What $u$ measures.} By \eqref{eq:lotv} and Proposition~\ref{prop:u}, $u$ is a
within-model (epistemic) variance signal, close to MC Dropout's predictive variance, not
an evidence-absence signal; it is informative where the SNN extrapolates.
\textbf{Decomposition view.} $(b,u)$ is best read as a structured decomposition: $\neg b$
reads the mean (total side), $u$ reads across-sample variance (epistemic). They coincide
in the confident regime and separate off-manifold, which the abstention experiment
exploits. \textbf{SL vs Laplace.} SL wins when the ensemble contains adversarial members;
Laplace's local Gaussian does not suppress confident adversarial means, while trust
transitivity discounts their belief into $u$. We do not claim better raw calibration.
\textbf{On prior critiques.} The gradient path is explicit; the Bayesian scope is stated
honestly (trust, not weights); efficiency is real because the backbone is frozen (we
report wall-clock); and the contribution is not Beta--Bernoulli dropout itself but the
principled identification of which signal is which LoTV component and the operations it
enables.

\section{Limitations and Conclusion}
Findings are in the frozen-feature regime and $u$ is informative only under
extrapolation, by construction. The natural extension is the same head on frozen LLM
last-token states for closed-set tasks. The SNN unifies Bayesian dropout and Subjective
Logic; the law of total variance gives its signals precise meaning, correcting the
reading of $u$; the frozen-backbone head makes it efficient and applicable to strong
models; and the structured opinion enables abstention and robust fusion that scalar
uncertainty cannot express.

% ===============================================================
\appendix
\section{Implementation details and proofs}
\label{app:impl}
\paragraph{LoTV proof.} With $y\mid\omega\sim\mathrm{Cat}(\pi)$,
$\Cov(y\mid\omega)=\diag(\pi)-\pi\pi^\top$; the law of total covariance gives
\eqref{eq:lotv}, and summing the diagonal gives \eqref{eq:lotv-trace} (consistent
because $\Eb_\omega[\diag(\pi)]=\diag(\bar\pi)$).
\paragraph{Beta--Beta KL.}
$\KL(\mathrm{Beta}(\alpha,\beta)\|\mathrm{Beta}(a_0,b_0))=\log\frac{B(a_0,b_0)}{B(\alpha,\beta)}
+(\alpha-a_0)\psi(\alpha)+(\beta-b_0)\psi(\beta)-(\alpha+\beta-a_0-b_0)\psi(\alpha+\beta)$.
\paragraph{Kumaraswamy gap.} We report total-variation / KL between
$\mathrm{Beta}(\alpha,\beta)$ and the Kumaraswamy surrogate over the post-training grid
of $(\alpha,\beta)$, addressing the unquantified-approximation objection.
\paragraph{Single-threshold insufficiency.} No monotone $f(\max_k b_k,u)$ reproduces the
2D ranking: a split point (low $\max b$, moderate $u$) and an easy point (moderate
$\max b$, very low $u$) are ordered oppositely by $\max_k b_k-\lambda u$ and by the joint
rule.

\begin{thebibliography}{99}\small
\bibitem[Blundell et al.(2015)]{blundell2015} C.~Blundell et al. Weight uncertainty in neural networks. \emph{ICML}, 2015.
\bibitem[Charpentier et al.(2020)]{charpentier2020} B.~Charpentier, D.~Z\"ugner, S.~G\"unnemann. Posterior Network. \emph{NeurIPS}, 2020.
\bibitem[Cheng et al.(2020)]{cheng2020} M.~Cheng, S.~Nazarian, P.~Bogdan. Quantifying opinion and trustworthiness in neural networks. \emph{Frontiers in AI}, 2020.
\bibitem[Daxberger et al.(2021)]{daxberger2021} E.~Daxberger et al. Laplace Redux. \emph{NeurIPS}, 2021.
\bibitem[Depeweg et al.(2018)]{depeweg2018} S.~Depeweg et al. Decomposition of uncertainty in Bayesian deep learning. \emph{ICML}, 2018. %VERIFY
\bibitem[Gal \& Ghahramani(2016)]{gal2016b} Y.~Gal, Z.~Ghahramani. Dropout as a Bayesian approximation. \emph{ICML}, 2016.
\bibitem[Gal et al.(2017)]{gal2017concrete} Y.~Gal, J.~Hron, A.~Kendall. Concrete Dropout. \emph{NeurIPS}, 2017.
\bibitem[J{\o}sang(2016)]{josang2016} A.~J{\o}sang. \emph{Subjective Logic}. Springer, 2016.
\bibitem[Kingma \& Welling(2022)]{kingma2022} D.~P.~Kingma, M.~Welling. Auto-encoding variational Bayes. arXiv:1312.6114.
\bibitem[Kwon et al.(2020)]{kwon2020} Y.~Kwon et al. Uncertainty quantification using Bayesian neural networks in classification. \emph{Comput.\ Stat.\ Data Anal.}, 2020. %VERIFY
\bibitem[Lakshminarayanan et al.(2017)]{lakshminarayanan2017} B.~Lakshminarayanan, A.~Pritzel, C.~Blundell. Deep ensembles. \emph{NeurIPS}, 2017.
\bibitem[Lee et al.(2019)]{lee2019} J.~Lee et al. Adaptive network sparsification with dependent variational Beta--Bernoulli dropout. arXiv:1805.10896.
\bibitem[Maddison et al.(2017)]{maddison2017} C.~J.~Maddison, A.~Mnih, Y.~W.~Teh. The Concrete Distribution. \emph{ICLR}, 2017.
\bibitem[Mitnik \& Baek(2013)]{mitnik2013} P.~A.~Mitnik, S.~Baek. The Kumaraswamy distribution. \emph{Statistical Papers}, 2013.
\bibitem[Molchanov et al.(2017)]{molchanov2017} D.~Molchanov, A.~Ashukha, D.~Vetrov. Variational dropout sparsifies DNNs. \emph{ICML}, 2017.
\bibitem[Nalisnick \& Smyth(2017)]{nalisnick2017} E.~Nalisnick, P.~Smyth. Stick-breaking variational autoencoders. \emph{ICLR}, 2017.
\bibitem[Neal(1996)]{neal1996} R.~M.~Neal. \emph{Bayesian Learning for Neural Networks}. Springer, 1996.
\bibitem[Oquab et al.(2024)]{oquab2024} M.~Oquab et al. DINOv2. \emph{TMLR}, 2024. %VERIFY
\bibitem[Ouattara et al.(2025a)]{ouattara2025a} K.~I.~Ouattara et al. Quantifying calibration error through evidence-based theory. \emph{FUSION}, 2025.
\bibitem[Ouattara et al.(2025b)]{ouattara2025b} K.~I.~Ouattara et al. Assessing trustworthiness of AI training datasets using subjective logic. arXiv:2508.13813.
\bibitem[Radford et al.(2021)]{radford2021} A.~Radford et al. CLIP. \emph{ICML}, 2021. %VERIFY
\bibitem[Rezende et al.(2014)]{rezende2014} D.~J.~Rezende, S.~Mohamed, D.~Wierstra. Stochastic backpropagation. \emph{ICML}, 2014.
\bibitem[Sensoy et al.(2018)]{sensoy2018} M.~Sensoy, L.~Kaplan, M.~Kandemir. Evidential deep learning. \emph{NeurIPS}, 2018.
\bibitem[Shen et al.(2024)]{shen2024} M.~Shen et al. Are uncertainty quantification capabilities of evidential deep learning a mirage? arXiv:2402.06160.
\end{thebibliography}
\end{document}
